The non-random assembly of functional motifs in  \nplant-pollinator networks

### 2022-04-12 Manuscript/Main_text/Main_text_Oikos.Rmd
Edits for Oikos submission

---
title: "The non-random assembly of functional motifs in plant-pollinator networks worldwide"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \floatplacement{figure}{H}
- \usepackage[skip=3pt]{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{lscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
output:
  pdf_document: default
  html_document: default
  word_document: default
always_allow_html: true
bibliography: bibliography.bib
biblio-style: apalike
csl: oikos.csl #oikos style from zotero styles
---

@@

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of mutualistic networks. Detecting those relationships is not trivial since they go beyond pairwise interactions that may get blurred when considering full network descriptors. Recent work has shown the network meso-scale can capture this important information. The meso-scale describes network subgraphs representing patterns of interactions between a small number of species (i.e., motifs) that constitute the building blocks of the whole network. Despite the possible implications of network motifs to better capture species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some motifs are consistently under- and over-represented worldwide, suggesting that the building blocks of plant-pollinator networks are not random. Furthermore, we find that distinct motif positions describing species ecological roles (e.g., generalization and number of indirect interactions) are occupied by different plant and floral visitor groups on both trophic levels. Bees appear consistently in generalist motif positions with low number of indirect interactions while the rest of floral visitor groups are associated with more specialised motif positions with higher number of indirect interactions. All plant groups are principally associated with specialised motif positions but only tall plant species with separated sexes (e.g., woody dioecious or monoecious species) are associated with lower number of indirect interactions. Interestingly, the realized combinations of different species groups within a motif can not be retrieved from their joint probability distributions, indicating that group combinations are not random either. Our result highlights the non-random structure of the meso-scale on plant-pollinators networks and the association of different plant and floral visitor groups with certain motifs that involve different ecological roles at a macro-ecological scale.

@@

Ecological communities are formed by a plethora of interacting species that form networks of interactions. Because of the tremendous complexity of these ecological networks, species interactions are generally condensed in metrics that summarize this information [@guimaraes2020]. Plant-pollinator interactions are no exception, and they are often studied with indices that aggregate the information at the species (micro-scale) or network level (macro-scale). For example, plant-pollinator network level approaches have identified common invariant structural properties across networks, including a degree distribution that decays as a power law [@jordano1987], nestedness [@bascompte2003], or modularity [@olesen2007]. In addition, the local species position within the network can define its degree of specialization [@bluthgen2006] or its role in connecting the rest of the community [@olesen2007]. Despite the unquestionable progress with the use of these metrics at both species and network level, condensing the information into a single metric implies the loss of relevant ecological information that obscures the understanding of species interactions [@cirtwill2018; @simmons2019].

Traditionally, plant-pollinator research has focused on direct interactions but overlooked indirect interactions (i.e., the mediated effect between two species by a third species), such as facilitative or competitive interactions between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Despite the widespread nature of indirect interactions in ecological communities [@strauss1991], plant-pollinator research often fails to finely capture those indirect interactions with the conventional analytical tools that condense the information either by species (e.g., interaction frequency) or in global topological indices (e.g., nestedness). Nonetheless, the emerging framework of network motifs (small sub-networks observed within a given network) in plant-pollinator research allows to consider both direct and indirect interactions [@simmons2019]. Motifs are often referred to as the building blocks of a network that depict subsets of interactions [@milo2002]. The analysis of network motifs in plant-pollinator networks have revealed that the different ecological processes that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect interactions [@simmons2020]. Yet, the global patterns of indirect interactions in real plant-pollinator networks are still unknown (e.g., over- and under- represented motifs).

By linking the structural properties of the meso-scale with the species' ecology we can advance in our understanding on the determinants of species interactions. For instance, different motifs can have different ecological meanings [@simmons2019] and the position within a motif can determine the species' functional role [@stouffer2012; @baker2015]. Yet, our understanding of the ecological implications of network motifs in real plant-pollinator networks remains in its infancy. Moreover, it is unclear how the species ecology and life-history traits determines its functional role within the network of interactions [@coux2016]. For example, large pollinators can forage larger distances [@greenleaf2007], deposit greater pollen quantities [@foldesi2021] and handle complex zygomorphic flowers in comparison with small pollinators that are restricted to lower floral complexity [@gong2009]. How this different pollinator behavior translates into their interaction topology is unknown. Similarly, recent empirical findings indicate that the meso-scale is the best descriptor of plant reproductive success [@allen2021], but little is known about how plants reproductive strategies shape their position within the network of interactions. Although some studies have evaluated plant reproductive strategies in plant-pollinator networks [@tur2013; @lazaro2020], this is an often overlooked aspect in a community context [@devaux2014] and rarely incorporated into plant-pollinator network studies. Hence, exploring how the main plant reproductive strategies integrate with the emergent motif framework could help to progress knowledge on plant-pollinator interactions.

@@

To investigate the structure of the meso-scale in plant-pollinator networks, we gathered published plant-pollinator networks with worldwide distribution. Then, we classified the plant and floral visitor species from these networks into groups that summarize their main life-history strategies. After, we calculated the normalized motif frequencies and investigated if these frequencies are similar or not to the ones expected by chance with the help of null models. Then, we also calculated the normalized group frequencies of plant and floral visitor for the different motif positions and investigated their overall representation by comparing these observed frequencies with their simulated counterparts. Finally, we tried to recover the observed motif combinations from the expected frequencies calculated as the product of the probabilities of the different plant and floral visitor groups within each motif.

@@

First, plant species were grouped through hierarchical clustering into the optimal number of functional groups that summarized the main plant reproductive strategies. For this, we used the trait dataset collated in @lanuza2021 that comprised a total of 1506 species including the 503 species considered in this study (**Table S2**). This dataset consisted on 8 floral, 4 reproductive and 3 vegetative traits excluding traits with high percentage of missing values (over 30$\%$; **Table S3**). We opted to calculate the plant functional groups on this larger set of species because of the higher accuracy when delimiting functional groups with that many variables and species [@dolnicar2014]. To feed the clustering analysis, we calculated the distance between the different qualitative and scaled quantitative variables with Gower distance [@gower1971]. For this, we used the function _gowdis_ with method _ward.D2_ from the package _FD_ version _1.0-12_ [@laliberte2014]. Finally, we conducted hierarchical clustering with the function _hclust_ from the _R_ stats package version _4.0.5_ and calculated the optimal number of clusters with the function _kgs_ from the package _maptree_ version _1.4-7_ [@white2009].

Second, floral visitors were grouped into taxonomic groups based on taxonomic rank as done similarly in other plant-pollinator studies [@fenster2004; @ollerton2009]. We opted to divide floral visitors on the taxonomic rank level and not with functional traits because the main orders of floral visitors differed in form and behaviour and had lower superior taxonomical complexity (i.e., floral visitors had 6 orders versus plants that had 38). Thus, this allowed us to group floral visitors into groups that represented approximately the main life strategies of the possible pollinators: (i) bees (Hymenoptera-Anthophila), (ii) non-bee Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Syrphidae-Diptera), (iv) non-syrphids-Diptera, (v) Lepidoptera and (vi) Coleoptera. However, a minor set of species belonged to other taxonomic groups that were considered in analyses but not discussed further because of their low representation in the full set of networks (3.55$\%$ of the total interactions recorded). These taxonomic groups were 'lizards', 'birds' and 'other insects'. This last group was formed by a mix of uncommon insect taxa on the full set of networks.

@@

Following previous work [@simmons2019; @simmons2020], we broke down the plant-pollinator networks into their constituent motifs that can capture both direct and indirect interactions. Prior to analyses, we turned the quantitative networks into qualitative (or binary) ones, where interactions are present or absent. All analyses were run considering singletons (species with only one interaction detected) but we conducted a second exploration without singletons (64.98$\%$ of interactions) to evaluate the effect of rare species on the observed patterns. Results without singletons are qualitatively consistent and are not further discussed in the main text (**Figures S2 and S3**).

We calculated the frequency of all motifs up to five nodes (17 different motifs in total; see **Figure 1**) for each empirical network, by using the _bmotif_ package version _2.0.2_ [@bmotifsimmons2019]. We focused exclusively on motifs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network motifs. Throughout the entire manuscript, floral visitors occupy the top nodes of the different motifs and plants the bottom ones. To control for variation in network size, motif frequencies were normalised as a proportion of the total number of motifs within each motif class (i.e., the number of nodes a motif contains). In addition, we excluded two-node motifs (or links) from our analyses because their normalised frequencies would always equal one. Like @simmons2020, we classified the different motifs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$). In general terms, plant and floral visitor groups on more densely connected motifs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected ones (e.g., medium-weak or weak).

```{r, echo=FALSE, message=FALSE, cache=FALSE, fig.pos="t", warning=FALSE,out.width="75%",fig.align="center", fig.cap="\\textbf{Figure 1.} (Adapted figure of Simmons et al., 2019). All possible motif network structures from two to five species in bipartite networks. There is a total of 17 possible motifs with 46 different positions denoted within each node. The different motifs are aggregated by their average path length classification wich goes from densely connected motifs (strong motifs coloured in yellow) to less connected motifs (weak motifs coloured in red).", fig.width=10,fig.height=8}
library(tidyverse)
library(RColorBrewer)
source("../../Scripts/Scripts_Alfonso/plot_motif_positionsV2.R")

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the _nullmodel_ function and the _vaznull_ model in the _bipartite_ package version _2.14_ [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. We discarded null models with more constraints (e.g., constraining for species degree) because the permutations did not generate sufficient variablity providing highly similar or identical networks to the original ones. After extracting the motif frequencies from the simulated networks, for each motif type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed motif frequencies. Motifs whose percentile is close to 0 or 100 are under- or over-represented in the empirical networks, respectively, and thus they cannot be predicted by connectance and the number of species alone. To summarize general patterns across networks, we used an intercept-only linear mixed model (LMM) per motif with the help of the package lmer version _1.1-21_ [@bates2015], where the response variable was the observed motif percentile per network. In these models, we used the study identifiers in **Table S1** as a random intercept. By doing so, we obtained estimates of the average motif frequency, but controlling for the variation at the study level.

@@

We calculated which plant and floral visitor groups were over- or under-represented in the different motif positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each plant and floral visitor group in a given network, we added the frequencies of those species that belong to the same group, and then, we normalised the resulting frequencies by dividing them by the total number of times that a corresponding group appears in any position within the same motif size class (i.e., number of species of a motif). Then, we calculated the percentile of the observed position frequencies for each group and network, just like we did with motif frequencies. To outline the general patterns of position frequencies across networks and groups, we fit a LMM per motif position, where the response variable was the observed position percentile per network. We used the group identifier as an explanatory variable and the study identifiers as a random intercept. Hence, we assessed the average motif frequency per species group, after controlling for the variation at the study level. Finally, we visualized with the help of the package _ComplexHeatmap_ version _2.6.2_ [@gu2016] over- and under-representation of plant and floral visitor groups on the different motif positions.

To understand if the different plant and floral visitor groups are associated with motif positions that represent different network roles we characterized the network roles of plant and floral visitor groups in terms of their specialization and number of indirect interactions. To that end, we calculated the specificity as a measure of specialization and we used the number of indirect interactions of the different motif positions to evaluate the association of these two node metrics and the group frequencies for the different motif positions. First, specificity ($s_i$) was calculated for each motif position following @poisot2015 as $s_i = L - l_i / L - 1$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given motif. Values of the vector s, range between 1, complete specialization (1 partner within the motif) and 0, complete generalism (all possible partners within the motif). Then, we explored graphically through linear models the association between the frequency of the different groups (response variable) and specificy (predictor). Second, we calculated the number of indirect interactions for each motif position by accounting only for the indirect effect of two species by a third species. That is, we explored exclusively indirect interactions within plants though shared floral visitors or within floral visitors though shared plants.

@@

Finally, we studied which combinations of plant and floral visitor groups tend to appear together within the same motifs (up to five nodes). That is, which combinations are over- or under-represented. This analysis used 57 out of the 60 networks available due to computational limitations to identify all the nodes in the motifs of the three networks with the highest number of links. To do so, for each of the 53,250 possible motif combinations, we estimated the observed and the expected probability of finding those combinations in empirical networks, respectively. Then, we determined whether the observed probabilities are likely to come from the expected probabilities or not. To calculate the observed probability ($p_i^O$) of the different groups within a motif $i$ (e.g., motif 3, 'bee' + 'bee' + 'selfing herbs'), we divided the number of times that $i$ appears in our set of empirical networks ($n_i^O$) by the sum of the number of times that each possible combination appeared: $p_i^O = n_i^O / \sum_{k=1}^{53,250} n_k^O$. To estimate the expected probability of a given motif combination $i$, $p_i^E$, firstly, we calculated the probability of finding a given group $x$ in the position $\alpha$ of $i$, $p_i(x,\alpha)$. Then, by assuming the independency of $p_i(x,\alpha)$, we computed the expected probability of the combination $i$ as the product of the probability of its pairs $(x,\alpha)$, that is, $p_i^E = \prod _{(x,\alpha )\in i} p_i(x,\alpha)$. To obtain $p_i(x,\alpha)$, we proceeded as follows. First, we calculated the number of times that the group $x$ appears in the position $\alpha$ in each empirical network $\eta$, denoted as $n_i^{\eta}(x,\alpha)$. Then, by controlling for the variation at the network level, we assessed the average value of the absolute frequency with which group $x$ appears in the position $\alpha$, $n_i(x,\alpha) = E\left [ n_i^{\eta}(x,\alpha) \right ]$, and estimated $p_i(x,\alpha)$ as $p_i(x,\alpha) = n_i(x,\alpha)/\sum_{k}n_i(k,\alpha)$. To obtain $n_i(x,\alpha)$, we fitted a LMM per motif position, where the response variable was the number of times that a given position was observed per network, the explanatory variable was the group identifier, and the random intercept was given by network identifiers nested within the study identifiers.

@@

The comparison of the plant and floral visitor group frequencies per motif position between empirical and simulated networks showed a `r over`$\%$ and `r under`$\%$ of over- (>75$^{th}$) and under-represented (<25th) groups in the different positions, respectively (**Figure 3**). Floral visitors showed a total of `r poll_over_75`$\%$ and `r poll_under_25`$\%$ of over- and under-represented groups in the different positions and plants `r plant_over_75`$\%$ and `r plant_under_25`$\%$ of over- and under-represented groups, respectively. Notably, the differences across groups were more marked for floral visitors than for plant groups (the differences between min and max percentiles per position were generally two-three times larger for floral visitors). From most over- to under-represented floral visitors groups on the different motif positions (indicated with the dendrogram order in **Figure 3**), we found: bees, non-syrphids Diptera, syrphids, Coleoptera, non-bee Hymenoptera and Lepidoptera. Although plant functional groups showed less differences between them, there were also more represented groups than others, thus from most over- to under-represented groups on the different motif positions we found: self incompatible perennials with large flowers, small outcrossing perennials, tall plants with unisexual flowers, selfing herbs and short lived outcrossers with long zygomorphic flowers.

>>>>

---
title: "The non-random assembly of functional motifs in  \nplant-pollinator networks"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \floatplacement{figure}{H}
- \usepackage[skip=3pt]{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{lscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[left]{lineno}
- \linenumbers

output:
  pdf_document: default
  html_document: default
  word_document: default
always_allow_html: true
bibliography: bibliography_no_doi.bib
biblio-style: apalike
csl: oikos.csl #funct. ecol. style from zotero styles
---

@@

Ecological processes leave distinct structural imprints on the species interactions that shape the topology of mutualistic networks. Detecting those relationships is not trivial since they go beyond pairwise interactions, but may get blurred when considering full network descriptors. Recent work has shown that the network meso-scale can capture this important information. The meso-scale describes network subgraphs representing patterns of interactions between a small number of species (i.e., motifs) that constitute the building blocks of the whole network. Despite the possible implications of network motifs to better capture species interactions, they remain overlooked in natural plant-pollinator networks. By exploring 60 empirical plant-pollinator networks from 18 different studies with wide geographical coverage we show that some motifs are consistently under- or over-represented worldwide, suggesting that the building blocks of plant-pollinator networks are not random. Furthermore, we find that distinct motif positions describing species ecological roles (e.g., generalisation and number of indirect interactions) are occupied by different plant and floral visitor groups on both trophic levels. Bees appear less frequently in specialised motif positions with high number of indirect interactions, while the rest of floral visitor groups are infrequent in generalised motif positions with low number of indirect interactions. All plant groups tend to be over-represented on specialised motif positions, except tall plant species with separated sexes (e.g., woody dioecious or monoecious species), which are more frequent on motif positions with low number of indirect interactions. Interestingly, the realized combinations of different species groups within a motif can not be retrieved from their joint probability distributions, indicating that group combinations are not random either. Our result highlights the non-random structure of the meso-scale on plant-pollinators networks and the association of different plant and floral visitor groups with certain motifs that involve different ecological roles at a macro-ecological scale.

@@

Ecological communities are formed by a plethora of interacting species that form networks of interactions. Because of the tremendous complexity of these ecological networks, species interactions are generally condensed in metrics that summarize this information [@guimaraes2020]. Plant-pollinator interactions are no exception, and they are often studied with indices that aggregate the information at the network (macro-scale) and species level (micro-scale). For example, plant-pollinator network level approaches have identified common invariant structural properties across networks, including a degree distribution that decays as a power law [@jordano1987], nestedness [@bascompte2003], or modularity [@olesen2007]. In addition, the local species position within the network can define its degree of specialisation [@bluthgen2006] or its role in connecting the rest of the community [@olesen2007]. Despite the unquestionable progress with the use of these metrics at both network and species level, condensing complex information into a single metric implies the loss of relevant ecological information that obscures the understanding of species interactions [@cirtwill2018; @simmons2019].

Traditionally, plant-pollinator research has focused on direct interactions but overlooked indirect interactions (i.e., the mediated effect between two species by a third species), such as facilitative or competitive interactions between plants for pollinators [@moeller2004; @sargent2008; @carvalheiro2014]. Despite the widespread nature of indirect interactions in ecological communities [@strauss1991], plant-pollinator research often fails to finely capture those indirect interactions with the conventional analytical tools that condense the information either by species (e.g., degree) or in global topological indices (e.g., nestedness). Nonetheless, the emerging framework of network motifs [small sub-networks observed within a given network which are often referred as its building blocks\; @milo2002] allows to consider both direct and indirect interactions [@simmons2019]. The analysis of network motifs in plant-pollinator networks have revealed that the different ecological processes that govern species interactions (e.g., species abundances versus trait-matching) can lead to different patterns of indirect interactions [@simmons2020]. Yet, the global patterns of indirect interactions in real plant-pollinator networks are still unknown (e.g., over- and under- represented motifs).

By linking the structural properties of the meso-scale with the species' ecology we can advance in our understanding on the determinants of species interactions. For instance, different motifs can have different ecological meanings [@simmons2019] and the position within a motif can determine the species' functional role [@stouffer2012; @baker2015]. Yet, our understanding of the ecological implications of network motifs in real plant-pollinator networks remains in its infancy. Moreover, it is unclear how the species ecology and life-history traits determine its functional role within the network of interactions [@coux2016]. For example, large pollinators can forage larger distances [@greenleaf2007], deposit greater pollen quantities [@foldesi2021] and handle complex zygomorphic flowers in comparison with small pollinators that are restricted to lower floral complexity [@gong2009]. Despite these obvious differences in pollinator behaviour, how contrasting species' ecology translates into different interaction patterns is still unknown. Similarly, recent empirical findings indicate that the meso-scale is the best descriptor of plant reproductive success [@allen2021], but little is known about how plants reproductive strategies shape their position within the network of interactions. Although some studies have evaluated plant reproductive strategies in plant-pollinator networks [@tur2013; @lazaro2020], this is an often overlooked aspect in a community context [@devaux2014] and rarely incorporated into plant-pollinator network studies. Hence, exploring how the main plant and floral visitor life-history strategies integrate with the emergent framework of network motifs could help to progress knowledge on the ecological processes that shape plant-pollinator interactions.

@@

To investigate the structure of the meso-scale in plant-pollinator networks, we gathered published plant-pollinator networks with worldwide distribution. Then, we classified the different plant and floral visitor species from these networks into groups that summarise their main life-history strategies. After, we calculated the normalised motif frequencies and investigated if these frequencies are similar or not to the ones expected by chance with the help of null models. Then, for each group of plants and floral visitors we investigated the ecological implications of their overall representation on the different motif positions, by comparing, on the one hand, the observed (normalised) frequencies of each group in the different motif positions with the frequencies of their simulated counterparts, and, on the other hand, the differences among the patterns of connections of motif positions. Finally, we compared if there are differences in the observed frequencies of motif combinations from the empirical networks and the expected frequencies of motif combinations calculated as the product of the probabilities of the different plant and floral visitor groups within the motif.

@@

First, plant species were grouped through hierarchical clustering into the optimal number of functional groups that summarized the main plant reproductive strategies. For this, we used the trait dataset collated in @lanuza2021 that comprised a total of 1506 species including the 503 species considered in this study (Table S2). This dataset consisted on 8 floral, 4 reproductive and 3 vegetative traits excluding traits with high percentage of missing values (over 30$\%$; Table S3). We opted to calculate the plant functional groups on this larger set of species because of the higher accuracy when delimiting functional groups with that many variables and species [@dolnicar2014]. To feed the clustering analysis, we calculated the distance between the different qualitative and scaled quantitative variables with Gower distance [@gower1971]. For this, we used the function _gowdis_ with method _ward.D2_ from the _FD_ package version _1.0-12_ [@laliberte2014]. Finally, we conducted hierarchical clustering with the function _hclust_ from the _R_ stats package version _4.0.5_ and calculated the optimal number of clusters with the function _kgs_ from the _maptree_ package version _1.4-7_ [@white2009].

Second, floral visitors were aggregated into taxonomic groups based on taxonomic rank as done similarly in other plant-pollinator studies [@fenster2004; @ollerton2009]. We opted to divide floral visitors on the taxonomic rank level and not with functional traits because the main orders of floral visitors differed in form and behaviour and had lower superior taxonomic complexity (i.e., floral visitors had 6 orders versus plants that had 38). Thus, this allowed us to group floral visitors into groups that represented approximately the main life strategies of the possible pollinators: (i) bees (Hymenoptera-Anthophila), (ii) non-bee Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Syrphidae-Diptera), (iv) non-syrphids-Diptera, (v) Lepidoptera and (vi) Coleoptera. However, a minor set of species belonged to other taxonomic groups that were considered in analyses but not discussed further because of their low representation in the full set of networks (3.55$\%$ of the total interactions recorded). These taxonomic groups were 'lizards', 'birds' and 'other insects'. This last group was formed by a mix of uncommon insect taxa on the full set of networks.

@@

Following previous work [@simmons2019; @simmons2020], we broke down the plant-pollinator networks into their constituent motifs that can capture both direct and indirect interactions. Prior to analyses, we turned the quantitative networks into qualitative (or binary) ones, where interactions are present or absent. This approach was preferred since quantitative motif analysis have been developed just recently and more work is needed to fully understand and interpret the information that they can convey [@simmons2020]. All analyses were run considering singletons (species with only one interaction detected) but we conducted a second exploration without singletons (64.98$\%$ of interactions) to evaluate the effect of rare species on the observed patterns. Results without singletons are qualitatively consistent and are not further discussed in the main text (Figs S2 and S3).

We calculated the frequency of all motifs up to five nodes (17 different motifs in total; see Fig. 1) for each empirical network, by using the _bmotif_ package version _2.0.2_ [@bmotifsimmons2019]. We focused exclusively on motifs up to five nodes in our analyses for computational reasons and for the ease of visualization and interpretation of a reduced set of less complex network motifs. Throughout the entire manuscript, floral visitors occupy the top nodes of the different motifs and plants the bottom ones. To control for variation in network size, motif frequencies were normalised as a proportion of the total number of motifs within each motif class (i.e., 2-species motifs, 3-species motifs, 4-species motifs and 5-species motifs). In addition, we excluded 2-species (or nodes) motifs from our analyses because its normalised frequencies would always equal one. Note that the aim is not comparing different networks, which may differ in sampling effort, but correctly describing the structure of each individual network. Like @simmons2020, we classified the different motifs by their average path length (mean number of links between all pairs of nodes) into four different groups: (i) complete or strong ($\overline{x} = 1.38$), (ii) fan or medium-strong ($\overline{x} = 1.48$), (iii) assymmetric complete or medium-weak ($\overline{x} = 1.60$) and (iv) core-peripheral or weak ($\overline{x} = 1.85$). In general terms, plant and floral visitor groups on more densely connected motifs (e.g., complete or fan) will experience lower number of indirect interactions in contrast to groups on less densely connected ones (e.g., medium-weak or weak).

```{r, echo=FALSE, message=FALSE, cache=FALSE, fig.pos="t", warning=FALSE,out.width="75%",fig.align="center", fig.cap="\\textbf{Figure 1.} Adapted figure of Simmons et al., (2019). All possible motif network structures from two to five species in bipartite networks grouped by the different motif classes (2-species, 3-species, 4-species and 5-species). There is a total of 17 possible motifs with 46 different positions denoted within each node. The different motifs are aggregated by their average path length classification wich goes from densely connected motifs (strong motifs coloured in yellow) to less connected motifs (weak motifs coloured in red). Note that two node motifs are excluded from this classification.", fig.width=10,fig.height=8}
library(tidyverse)
library(RColorBrewer)
source("../../Scripts/Scripts_Alfonso/plot_motif_positionsV2.R")

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the _nullmodel_ function and the _vaznull_ model in the _bipartite_ package version _2.14_ [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. We discarded null models with more constraints (e.g., constraining for species degree) because the permutations did not generate sufficient variablity and provided highly similar or identical networks to the original ones. After extracting the motif frequencies from the simulated networks, for each motif type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed motif frequencies. Motifs whose percentile is close to 0 or 100 are under- or over-represented in the empirical networks, respectively, and thus they cannot be predicted by connectance and the number of species alone. To summarize general patterns across networks, we used an intercept-only linear mixed model (LMM) per motif with the help of the _lme4_ package version _1.1-21_ [@bates2015], where the response variable was the observed motif percentile per network. Note that as we are only interested in modelling the intercept value, no predictors were specified. In these models, we used the study identifiers in Table S1 as a random intercept. By doing so, we obtained estimates of the average motif frequency, but controlling for the variation at the study level.

@@

We calculated which plant and floral visitor groups were over- or under-represented in the different motif positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each plant and floral visitor group in a given network, we added the frequencies of those species that belong to the same group, and then, we normalised the resulting frequencies by dividing them by the total number of times that a corresponding group appears in any position within the same motif size class (i.e., 2, 3, 4 or 5 species motifs). Then, we calculated the percentile of the observed position frequencies for each group and network, just like we did with motif frequencies. To outline the general patterns of position frequencies across networks and groups, we fit a LMM per motif position, where the response variable was the observed position percentile per network. We used the group identifier as an explanatory variable and the study identifiers as a random intercept. Hence, we assessed the average motif frequency per species group, after controlling for the variation at the study level. Finally, we visualized with the help of the _ComplexHeatmap_ package version _2.6.2_ [@gu2016] over- and under-representation of plant and floral visitor groups on the different motif positions.

To understand if the different plant and floral visitor groups are associated with motif positions that represent different network roles, we characterized the roles of each motif positions for plant and floral visitor groups in terms of their specialisation and number of indirect interactions. To that end, we calculated the specificity of each motif position following @poisot2015 as $s_i = L - l_i / L - 1$, where $L$ is the total number of groups and $l_i$ the number of interaction partners of the group $i$ in a given motif. Values of the vector $si$, range between 1, complete specialisation (1 partner within the motif) and 0 complete generalism (all possible partners within the motif). Then, because all nodes within a motif are connected, we calculated the number of indirect interactions ($n_i$) for each motif position and trophic guild (i.e., plants or floral visitors) as $n_i=z_i-1$, where $z_i$ is the total number of nodes for a trophic guild in a given motif. Thus, we considered exclusively indirect interactions within plants though shared floral visitors or within floral visitors though shared plants. Finally, we analysed in a single model how specialization and the number of indirect interactions were associated with the observed percentiles for each trophic guild and motif position (see paragraph above). For this, we fitted a LMM model with the help again of the _lme4_ package where we considered specialization and number of indirect interactions as response variables with the interaction by plant and floral visitor group for both response variables. Further, we considered as random factor motif position nested within study system. All model diagnostics were conducted with the help of the _Dharma_ package [@hartig2017] version _0.3.3.0_ and model outputs were visualized with the _visreg_ package [@breheny2020] version 2.7.0.

@@

Finally, we studied which combinations of plant and floral visitor groups tend to appear together within the same motifs (up to five nodes). That is, which combinations are over- or under-represented. This analysis used 57 out of the 60 networks available due to computational limitations to identify all the nodes in the motifs of the three networks with the highest number of links. To do so, for each of the 53,250 possible motif combinations, we estimated the observed and the expected probability of finding those combinations in empirical networks, respectively. Then, we determined whether the observed probabilities are likely to come from the expected probabilities or not. To calculate the observed probability ($p_i^O$) of the different groups within a motif $i$ (e.g., motif 3, 'bee' + 'bee' + 'selfing herbs'), we divided the number of times that $i$ appears in our set of empirical networks ($n_i^O$) by the sum of the number of times that each possible combination appeared: $p_i^O = n_i^O / \sum_{k=1}^{53,250} n_k^O$. To estimate the expected probability ($p_i^E$) of a given motif combination $i$, we assumed that the probability of finding a given group $x$ in the position $\alpha$ is independent of the probabilities associated with other positions and groups in that motif. Therefore, the expected probability ($p_i^E$) of the motif combination $i$ is given by the product of the probabilities of its constituents, i.e., $p_i^E = \prod _{(x,\alpha )\in i} p_i(x,\alpha)$, where $p_i(x,\alpha)$ is the probability of finding a given group $x$ in the position $\alpha$ of $i$. Then, by controlling for the variation at the network level with an LMM (see below), we calculated the average number of times that the group $x$ appears in the position $\alpha$ (denoted as $n_i(x,\alpha)$). To obtain the expected probability $p_i(x,\alpha)$ we divided the number of times that the group $x$ appears in the position $\alpha$ by the the total number of times that any group was observed in that position (that is, $p_i(x,\alpha) = n_i(x,\alpha)/\sum_{k}n_i(k,\alpha)$). The average number of times that the group $x$ appears in the position $\alpha$ ($n_i(x,\alpha)$) was calculated with a LMM for each position, where the response variable was the number of times that a group was observed at a given position in each real-world network, the explanatory variable was the group identifier, and the random intercept was given by network identifiers nested within the study identifiers.

@@

The comparison of the plant and floral visitor group frequencies per motif position between empirical and simulated networks showed a `r over`$\%$ and `r under`$\%$ of over- (>75$^{th}$) and under-represented (<25th) groups in the different positions, respectively (Fig. 3). Floral visitors showed a total of `r poll_over_75`$\%$ and `r poll_under_25`$\%$ of over- and under-represented groups in the different positions and plants `r plant_over_75`$\%$ and `r plant_under_25`$\%$ of over- and under-represented groups, respectively. Notably, the differences across groups were more marked for floral visitors than for plant groups (the differences between min and max percentiles per position were generally two-three times larger for floral visitors). From most over- to under-represented floral visitors groups on the different motif positions (indicated with the dendrogram order in Fig. 3), we found: bees, non-syrphids Diptera, syrphids, Coleoptera, non-bee Hymenoptera and Lepidoptera. Although plant functional groups showed less differences between them, there were also more represented groups than others, thus from most over- to under-represented groups on the different motif positions we found: self incompatible perennials with large flowers, small outcrossing perennials, tall plants with unisexual flowers, selfing herbs and short lived outcrossers with long zygomorphic flowers.

For both floral visitor and plant groups we found over- and under-representation on different motif positions that implied different ecological roles. First, for floral visitors, we found that bees were less frequent on motif positions that involved specialization (i.e., lower specificity values) while the rest of taxonomic groups (non-syrphid Diptera, syrphids, Lepidoptera, Coleoptera and non-bee Hymenoptera) were infrequent in generalised motif positions with low number of indirect interactions (Fig. S6 A). Although plants did not show differences with null models on their frequencies, all plant but tall plants with small unisexual flowers showed a slight tendency to be over-represented on specialised motif positions (Fig. S6 B). Second, regarding the number of indirect interactions, bees were over-represented in motif positions with low number of indirect interactions (Fig. S7 A) and the rest of floral visitor groups showed no differences across positions in comparison with simulated networks. For plants, all groups showed a slight trend to be over-represented on motif positions with higher number of indirect interactions (Fig. S7 B). When we further explore the observed frequencies by number of indirect interactions for floral visitor and plant groups, we found that both bees and tall plants with small unisexual flowers were the only groups with high frequencies on motif positions associated with low number interactions in comparison with the other floral visitor and plant groups (Figs S8 and S9).
